---
title: "hw05"
format: html
editor: visual
---

# Analyzing Parasites Killing Bee Colonies in Arizona From 2015 - 2021

## Intro

Two national datasets regarding bee populations ("colony.csv") and stressors causing bee population decline ("stressor.csv") were analyzed for the purposes of analyzing and predicting liner regression. Data in both datasets have been categorized based on season and the two main culprits leading to population decline are "Varroa mites" followed by "Other pests/parasites".

Varroa mites are parasites that feed off the bodily fluids of adult bees and younger, newborn bees within a bee population, leading to a significant decline in bee population across the entire nation. This assignment looks to analyze the linear affect that this parasite specifically has on the Arizona bee population from 2015 - 2021.

## Data Preprocessing: Import Packages & Two Datasets

```{r}

# Load packages

if (!require(pacman))
  install.packages("pacman")

pacman::p_load(tidymodels,
               dplyr,
               tidyverse,
               ranger,
               randomForest,
               glmnet,
               here,
               formattable,
               dlookr,
               yardstick,
               ggplot,
               gplots,
               gridExtra)

# Import two correlated datasets, "colony" and "stressor": 

# Read colony data (dataset is from 2015 - 2021)
colony <- read_csv(here('./data/colony.csv'))
arizona_colony <- filter(colony, state %in% "Arizona")
arizona_colony

# Read stressor data
stressor <- read_csv(here('./data/stressor.csv'))
arizona_stressor <- filter(stressor, state %in% "Arizona")
arizona_stressor
```

## Data Preprocessing: Cleaning "Colony" Dataset

```{r}

# Analyze "arizona_colony" data: 

arizona_colony |> summary()

# Check for missing values: 

na_values_arizona_colony <- colSums(is.na(arizona_colony))
na_values_arizona_colony

# Analyze and inspect the rows missing values: 

missing_rows_arizona_colony <- arizona_colony %>%
  filter(!complete.cases(.))
missing_rows_arizona_colony

# Remove "colony_reno_pct" (impute):

arizona_colony %>%
    select(-colony_reno_pct) %>%
    filter(!complete.cases(.))
arizona_colony

# Don't impute April-June 2019, and remove row, as this is only row with null values and entire row of data wasn't collected: 

arizona_colony %>% na.omit()
```

## Data Preprocessing: Analyzing "Stressors" and Major Stressors Causing Decline in AZ Bee Population

```{r}

# Filter stressor data: 

arizona_stressor |> summary()
arizona_stressor

# Rearrange stressor values from high to low:

arizona_stressor %>% arrange(desc(stress_pct))

# Analyze "stress_pct" distribution with a box plot:

boxplot(arizona_stressor$stress_pct, main = "Boxplot of stress_pct")

# "Varroa mites" are having the main effect on the declining AZ bee population from a few seasonal months in 2015-2019, followed by "Other pests/parasites" in April-June 2016, then back to "Varroa mites from a few seasonal months in 2016-2018.
```

## Data Preprocessing: Combine Both Datasets Based on Population Count and Stressors

```{r}

# First filter "Stressor" dataset for top 2 stressors then concatonate with corresponding bee population from "Colony" dataset

top_arizona_stressors <- arizona_stressor %>%
  filter(stressor %in% c("Varroa mites", "Other pests/parasites"))
top_arizona_stressors

combined_dataset <- merge(top_arizona_stressors, colony, by = c("year", "months", "state"))

combined_dataset %>% arrange(desc(stress_pct))

# Remove N/A values from combined_dataset:

combined_dataset <- na.omit(combined_dataset)
```

## Data Preprocessing: Finding Correlations Between Numeric Columns

```{r}

# Find numeric columns in combined dataset: 

combined_dataset |> summary()

# Need to remove non-numeric columns for pairs function:
numeric_data <- combined_dataset %>%
  select_if(is.numeric)

# Run combined_dataset in a correlation matrix:
cor_matrix <- cor(numeric_data)
cor_matrix

# Run pairs function on filtered dataset:
pairs(numeric_data)

# Create heatmap of numeric values:
heatmap(cor(numeric_dataset), col = colorRampPalette(c("blue", "white", "red"))(20))
```

## Linear Modeling

```{r}

# Creating regression model 1:
model1 <- lm(colony_n ~ colony_added, data = combined_dataset)
model1

# Creating regression model 2:

model2 <- lm(colony_n ~ colony_lost, data = combined_dataset)
model2

# Diagnostic plots for model1:
par(mfrow = c(2, 2))  
plot(model1)

# Diagnostic plots for model2:
par(mfrow = c(2, 2))  
plot(model2)
```

## Part I of II: Colony Lost Analysis

This dataset views the trend of bee colonies lost in Arizona over a six-year period (2015-2021).

```{r}

# Analyze Colony Lost - Max & Min 

stressor_counts <- table(stressor$stressor)
stressor_counts

max_colony_stressor <- max(stressor$stressor)
min_colony_stressor <- min(stressor$stressor)
max_colony
min_age
```

## Part II of II: Stressors Killing Bee Colonies

```{r}

# Need to classify "Stressors" as factors instead of characters:



my_data$Name <- as.factor(my_data$Name)
```
