---
title: "hw05"
format: html
editor: visual
---

# Analyzing Parasites Killing Bee Colonies in Arizona From 2015 - 2021

## Intro

Two national datasets regarding bee populations ("colony.csv") and stressors causing bee population decline ("stressor.csv") were analyzed for the purposes of analyzing and predicting liner regression. Data in both datasets have been categorized based on season and the two main culprits leading to population decline are "Varroa mites" followed by "Other pests/parasites".

Varroa mites are parasites that feed off the bodily fluids of adult bees and younger, newborn bees within a bee population, leading to a significant decline in bee population across the entire nation. This assignment looks to analyze the linear affect that this parasite specifically has on the Arizona bee population from 2015 - 2021.

## Data Preprocessing: Import Packages & Two Datasets

```{r}

# Load packages

if (!require(pacman))
  install.packages("pacman")

pacman::p_load(tidymodels,
               dplyr,
               tidyverse,
               ranger,
               randomForest,
               glmnet,
               here,
               formattable,
               dlookr,
               yardstick,
               ggplot,
               gplots,
               caret,
               gridExtra)

# Import two correlated datasets, "colony" and "stressor": 

# Read colony data (dataset is from 2015 - 2021)
colony <- read_csv(here('./data/colony.csv'))
arizona_colony <- filter(colony, state %in% "Arizona")
arizona_colony

# Read stressor data
stressor <- read_csv(here('./data/stressor.csv'))
arizona_stressor <- filter(stressor, state %in% "Arizona")
arizona_stressor
```

## Data Preprocessing: Cleaning "Colony" Dataset

```{r}

# Analyze "arizona_colony" data: 

arizona_colony |> summary()

# Check for missing values: 

na_values_arizona_colony <- colSums(is.na(arizona_colony))
na_values_arizona_colony

# Analyze and inspect the rows missing values: 

missing_rows_arizona_colony <- arizona_colony %>%
  filter(!complete.cases(.))
missing_rows_arizona_colony

# Remove "colony_reno_pct" (impute):

arizona_colony %>%
    select(-colony_reno_pct) %>%
    filter(!complete.cases(.))
arizona_colony

# Don't impute April-June 2019, and remove row, as this is only row with null values and entire row of data wasn't collected: 

arizona_colony %>% na.omit()
```

## Data Preprocessing: Analyzing "Stressors" and Major Stressors Causing Decline in AZ Bee Population

```{r}

# Filter stressor data: 

arizona_stressor |> summary()
arizona_stressor

# Rearrange stressor values from high to low:

arizona_stressor %>% arrange(desc(stress_pct))

# Analyze "stress_pct" distribution with a box plot:

boxplot(arizona_stressor$stress_pct, main = "Boxplot of stress_pct")

# "Varroa mites" are having the main effect on the declining AZ bee population from a few seasonal months in 2015-2019, followed by "Other pests/parasites" in April-June 2016, then back to "Varroa mites from a few seasonal months in 2016-2018.
```

## Data Preprocessing: Combine Both Datasets Based on Population Count and Stressors

```{r}

# First filter "Stressor" dataset for top 2 stressors then concatonate with corresponding bee population from "Colony" dataset

top_arizona_stressors <- arizona_stressor %>%
  filter(stressor %in% c("Varroa mites", "Other pests/parasites"))
top_arizona_stressors

combined_dataset <- merge(top_arizona_stressors, colony, by = c("year", "months", "state"))

combined_dataset %>% arrange(desc(stress_pct))

# Remove N/A values from combined_dataset:

combined_dataset <- na.omit(combined_dataset)
```

## Data Preprocessing: Finding Correlations Between Numeric Columns

```{r}

# Find numeric columns in combined dataset: 

combined_dataset |> summary()

# Need to remove non-numeric columns for pairs function:
numeric_data <- combined_dataset %>%
  select_if(is.numeric)

# Run combined_dataset in a correlation matrix:
cor_matrix <- cor(numeric_data)
cor_matrix

# Run pairs function on filtered dataset:
pairs(numeric_data)

# Create heatmap of numeric values:
heatmap(cor(numeric_dataset), col = colorRampPalette(c("blue", "white", "red"))(20))
```

## Designing & Creating Linear Models

```{r}

# Creating regression model 1:
model1 <- lm(colony_n ~ colony_added, data = combined_dataset)
model1

# Creating regression model 2:

model2 <- lm(colony_n ~ colony_lost, data = combined_dataset)
model2

# Diagnostic plots for model1:
par(mfrow = c(2, 2))  
plot(model1)

# Diagnostic plots for model2:
par(mfrow = c(2, 2))  
plot(model2)
```

## Testing Linear Model

```{r}

# Set seed:
set.seed(123)

# Conduct 80% training/20% testing split:
index <- createDataPartition(combined_dataset$colony_n, p = 0.8, list = FALSE)
train_data <- combined_dataset[index, ]
test_data <- combined_dataset[-index, ]

# Make predictions on the testing set
predictions <- predict(model2, newdata = test_data)

# Evaluate the model performance on the testing set (you can use different metrics)
# For example, you might want to use mean squared error (MSE)
mse <- mean((predictions - test_data$colony_n)^2)
print(paste("Mean Squared Error on Test Set:", mse))
```

## Conclusion

The methodologies and data mining techniques used for linear regression analysis in this report conclude insignificant and high unreliable findings when trying to predict the relationship between parasites and their effects on the decline of bee population. This indicates that the combined datasets needs to be re-evaluated and better filtered and imputed during the correlation analysis of the Exploratory Data Analysis process.

But most importantly, additional linear models must be chosen and used in cross-validation testing to determine the best model that will better fit the data than the singular model currently used in this analysis, which should therefore better predict the future Arizona bee population and how much the Varroa mite and other parasites may affect this population.
